Perturbing SU($n$)-symmetric interactions

### 2020-03-09 collective_sun/sun_perturbation.tex
simplify section on using translational invariance to simplify the multibody eigenvalue problem

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  In order to do
so, we label spins $p,q$ by the position of the corresponding lattice
site in $\ZZ_L^D$, and use translational invariance of the couplings
$\m h$ to expand
\begin{align}
  \m h \equiv \sum_{p,q\in\ZZ_L^D} h_{pq} \op{p}{q}
  = \sum_{q,d\in\ZZ_L^D} h_{q+d,q} \op{q+d}{q}
  = \sum_{d\in\ZZ_L^D} h_{d,0} \m s_d,
  &&
  \m s_d \equiv \sum_{q\in\ZZ_L^D} \op{q+d}{d}.
\end{align}
Starting from this form of $\m h$ and the expansion of $\m w$ in
\eqref{eq:sym_tensor}, we can work out that
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{d\in\ZZ_L^D} \sum_{a=1}^M w_{k_{d;a}}
  h_{d,0} w_{k_{d;a}} \m m_k^{\t{TS}},
\end{align}
where $k_{d;a}$ denotes the spins acquired by displacing the $a$-th
spin of $k$ by $d$.  With a bit of additional work, we can also
simplify
\begin{align}
  \tilde{\m h}
  = \sum_{k\in\T_{LD}\p{M}} \sp{M h_0 - \sum_{p,q\in k} h_{pq}}
  \m m_k^{\t{TS}},
  &&
  h_0 \equiv \sum_{d\in\ZZ_L^D} h_{d,0},
\end{align}
which implies that
\begin{align}
  \tilde{\m h} * \m w = M h_0 \m w
  - \sum_{k\in\T_{LD}\p{M}} \sum_{p,q\in k} h_{pq} w_k \m m_k^{\t{TS}}.
\end{align}
Putting everything together, the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} can be written in the form
\begin{align}
  \sum_{k\in\T_{LD}\p{M}} \sp{\sum_{p,q\in k} h_{pq} w_k
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M w_{k_{d;a}}}
  \m m_k^{\t{TS}}
  = \p{M h_0 + \Delta} \m w,
\end{align}
or equivalently
\begin{align}
  \check{\m h} \c \v{\m w} = \p{M h_0 + \Delta} \v{\m w},
  \label{eq:H_VM_eig_simp}
\end{align}
where
\begin{align}
  \check{\m h} \equiv
  \sum_{k\in\T_{LD}\p{M}} \ket{k}
  \sp{\sum_{p,q\in k} h_{pq} \bra{k}
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M \bra{k_{d;a}}},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k},
\end{align}
and, to keep notation simple, we define $\ket{k}=0$ for all $k$ with
repeated values, and enforce $\ket{j}=\ket{k}$ if spins $j$ and spins
$k$ are the same up to a global translation.  The matrix
$\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Additional symmetries}

If the interaction Hamiltonian $H_0$ and multi-body perturbation
$\V_M$ have additional symmetries, such as isotropy or nonchirality,
then we can use this symmetry to reduce the eigenvalue problem in
\eqref{eq:H_VM_eig_simp} even further.  To this end, we define the set
$\S_{LD}\p{M}$ of equivalence classes in $\T_{LD}\p{M}$ under the
relevant symmetry.  In the case of isotropic two-body perturbations,
for example, the set of translationally-symmetrized spin choices
$\T_{LD}\p{2}$ can be classified by {\it displacements} on a lattice
(i.e.~a vector from one chosen spin to the other), and we can equate
all displacements that have the same magnitude, so elements of
$\S_{LD}\p{2}$ can be classified by all by {\it distances} between any
two points on the lattice.  For any member $s\in\S_{LD}\p{M}$, we then
define a uniform superposition $\ket{s}$ over all corresponding
members of $\T_{LD}\p{M}$, as well as the projector $\P_\S$ onto the
space of uniform superpositions:
\begin{align}
  \ket{s} \equiv \f1{\sqrt{\abs{s}}} \sum_{k\in s} \ket{k},
  &&
  \P_\S \equiv \sum_{s\in\S_{LD}\p{M}} \op{s},
\end{align}
where $\abs{s}$ is the number of members of $\T_{LD}\p{M}$ in the
equivalence class $s$.  We then expand the (symmetric) coefficient
vector
\begin{align}
  \v{\m w} = \P_\S \v{\m w}
  = \sum_{s\in\S_{LD}\p{M}} \ut{w}_s \ket{s},
  &&
  \ut{w}_s \equiv w_s \sqrt{\abs{s}},
\end{align}
as well as the symmetric projection of the matrix $\check{\m h}$:
\begin{align}
  \P_\S \check{\m h} \P_\S
  = \sum_{s\in\S_{LD}\p{M}} \ket{s} \sp{\f{h_s}{\abs{s}} \bra{s}
    + \sum_{d\in\ZZ_L^D} \f{h_{d,0}}{\sqrt{\abs{s}}} \sum_{a=1}^M
    \f1{\sqrt{\abs{s_{d;a}}}} \bra{s_{d;a}}},
\end{align}
where $h_s\equiv\sum_{p,q\in k}h_{pq}$ for any $k\in s$; and,
enforcing a canonical choice of $k\in s$ for all $s\in\S_{LD}\p{M}$,
$s_{d;a}$ is the equivalence class of the corresponding $k_{d;a}$,
with $\ket{s_{d;a}}\equiv0$ if $\ket{k_{d;a}}=0$.  The
symmetry-reduced eigenvalue problem is then
\begin{align}
  \P_\S \check{\m h} \P_\S \c \v{\m w} = \p{M h_0 + \Delta} \v{\m w}.
\end{align}

>>>>

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  We therefore
expand
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} w_{k_{ap}} \m m_k^{\t{TS}},
  &&
  \tilde{\m h} * \m w = \sum_{k\in\T_{LD}\p{M}}
  \tilde h_k w_k \m m_k^{\t{TS}},
\end{align}
which implies that the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} can be written as
\begin{align}
  \check{\m h}\c\v{\m w} = \Delta\v{\m w},
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} \op{k}{k_{ap}}
  - \sum_{k\in\T_{LD}\p{M}} \tilde h_k \op{k},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k},
\end{align}
and, by slight abuse of notation we identify an equivalence classes
$k\in\T_{LD}\p{M}$ with a canonical choice of spins
$\tilde k\in k\subset\C_N\p{M}$ (such that $k_{ap}=\tilde k_{ap}$,
with $\tilde k_{ap}$ equal to $\tilde k$ with the $a$-th spin replaced
by $p$); and we define $\ket{j}=0$ for all $j\in\C_N\p{M}$ with
repeated values.  The matrix $\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.
